---
title: "3C 279: Modelling Brightness and Variability Before vs. After the 2023 Dimming"
subtitle: "A mixed-effects (GLMM/LMM) analysis of AAVSO magnitudes"
author: "AAVSO — analysis for Rodney H"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    fig-width: 8
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
# Packages -------------------------------------------------------------------
library(tidyverse)
library(lubridate)
library(scales)
library(patchwork)
library(lme4)      # linear mixed model (mean structure)
library(glmmTMB)   # location-scale model: lets us model the *variance* too
library(broom.mixed)
library(knitr)

# Project root ---------------------------------------------------------------
# All paths are relative to this .qmd so the report is portable.
WD       <- getwd()
DATA_CSV <- file.path(WD, "rawdata_279-3C_cleaned.csv")

# Epoch boundaries -----------------------------------------------------------
# The scientific question: has variability changed since the 2023 dimming?
# We compare the "pre" era (before 2022) with the "dimming" era (2023 onward),
# treating 2022 as a transition year that is EXCLUDED from the contrast so the
# two groups are cleanly separated.
PRE_END      <- ymd("2022-01-01")  # observations before this date = "pre_2022"
DIMMING_FROM <- ymd("2023-01-01")  # observations on/after this date = "dimming_2023on"
```

```{r utilities}
# Utility functions ----------------------------------------------------------

#' Julian Date -> calendar Date (UTC).
#' JD 2440587.5 corresponds to 1970-01-01 00:00 UTC (the Unix epoch).
jd_to_date <- function(jd) as.Date(jd - 2440587.5, origin = "1970-01-01")

#' Standard error of the mean
se <- function(x) sqrt(var(x, na.rm = TRUE) / sum(!is.na(x)))

#' +/-3 SD summary (kept from the AAVSO template for a familiar readout)
sd3 <- function(x) {
  m <- round(mean(x, na.rm = TRUE), 3)
  s <- round(sd(x,   na.rm = TRUE), 3)
  tibble(
    label = c("-3sd", "-2sd", "-1sd", "Mean", "+1sd", "+2sd", "+3sd"),
    value = c(m - 3*s, m - 2*s, m - s, m, m + s, m + 2*s, m + 3*s)
  )
}
```

# Abstract {.unnumbered}

::: {.callout-note appearance="simple"}
**In plain language.** This report asks whether the quasar **3C 279** has really
dimmed since 2023, and whether its brightness has become steadier or more
erratic — using AAVSO amateur observations going back to about 1999.

The answer to both is **yes, it changed — it is now fainter *and* quieter.**
Comparing the years before 2022 with 2023 onward (2022 is left out as a
transition), and looking at the well-standardised Johnson **V** filter so we
compare like with like, 3C 279 is about **1.7 magnitudes fainter** than before
(a magnitude is a factor of ~2.5 in brightness, so this is a large drop), and
its night-to-night scatter has roughly **halved**. In other words, the object
didn't just dim — it also settled into a calmer, less variable state.

One important subtlety, spelled out in the analysis: if you lump *all* filters
together, the "quieter" signal almost disappears. That is not because the source
stayed variable — it is because different filters were used in different eras and
they have different natural scatter. Once we hold the filter fixed (the V-only
check), the drop in variability is clear. This is exactly the kind of confound
the mixed-effects model is designed to handle.

*Everything below shows the data cleaning, the statistical models, and the
diagnostic checks behind these two conclusions.*
:::

# Data

## Load and clean

The raw file has columns `JD`, `mag`, `I_Magnitude`, `band`, `val`.

- `I_Magnitude` is constant (all `0`) and carries no information — dropped.
- `val` is a free-text note field ("Z", "V", exposure strings) — not used in the model.
- `band` is the photometric passband. **Magnitudes are not comparable across
  bands**, so band is a first-class effect in the model, and non-photometric /
  junk codes (`N/A`, `TG`, `SR`, `CV`) are removed.
- Rows with a missing `mag` are dropped.

```{r load-clean}
raw <- read_csv(DATA_CSV, show_col_types = FALSE)

# Photometric bands we trust for a magnitude comparison.
KEEP_BANDS <- c("V", "CR", "B", "R", "I", "Vis.")

dat <- raw |>
  transmute(
    jd    = as.numeric(JD),
    mag   = as.numeric(mag),
    band  = as.character(band)
  ) |>
  filter(!is.na(jd), !is.na(mag), band %in% KEEP_BANDS) |>
  mutate(
    date  = jd_to_date(jd),
    year  = year(date),
    band  = factor(band, levels = KEEP_BANDS)
  ) |>
  # Guard against obviously bad magnitudes (3C 279 lives ~11-18 mag).
  filter(mag > 8, mag < 20) |>
  arrange(date)

cat(sprintf("Rows: %d raw -> %d after cleaning\n", nrow(raw), nrow(dat)))
glimpse(dat)
```

## Define the epoch contrast

```{r epochs}
model_dat <- dat |>
  mutate(
    epoch = case_when(
      date <  PRE_END      ~ "pre_2022",
      date >= DIMMING_FROM ~ "dimming_2023on",
      TRUE                 ~ NA_character_        # 2022 transition -> excluded
    )
  ) |>
  filter(!is.na(epoch)) |>
  mutate(epoch = factor(epoch, levels = c("pre_2022", "dimming_2023on")))

# How much data, and which bands, land in each epoch?
model_dat |>
  count(epoch, band) |>
  pivot_wider(names_from = band, values_from = n, values_fill = 0) |>
  kable(caption = "Observation counts by epoch and band (2022 excluded as transition)")

model_dat |>
  group_by(epoch) |>
  summarise(
    n         = n(),
    mean_mag  = mean(mag),
    sd_mag    = sd(mag),
    var_mag   = var(mag),
    .groups   = "drop"
  ) |>
  kable(digits = 3,
        caption = "Brightness (mean) and variability (SD/variance) by epoch")
```

> Reminder on magnitudes: **larger `mag` = fainter**. If the object has dimmed
> since 2023, we expect a *higher* mean magnitude in the `dimming_2023on` epoch.

\newpage

# The light curve

```{r light-curve, fig.width=10, fig.height=5}
p_lc <- ggplot(dat, aes(x = date, y = mag, colour = band)) +
  geom_point(alpha = 0.35, size = 0.9) +
  # Shade the excluded 2022 transition band.
  annotate("rect", xmin = PRE_END, xmax = DIMMING_FROM,
           ymin = -Inf, ymax = Inf, alpha = 0.08, fill = "grey20") +
  geom_vline(xintercept = c(PRE_END, DIMMING_FROM),
             linetype = "dashed", colour = "grey40") +
  scale_y_reverse() +                       # brighter (smaller mag) at the top
  labs(
    title    = "3C 279 — AAVSO light curve with epoch split",
    subtitle = "Shaded band = 2022 transition (excluded from the contrast)",
    x = NULL, y = "Magnitude (fainter →)", colour = "Band"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

p_lc
```

# Distribution by epoch

```{r dist-by-epoch, fig.width=9, fig.height=4.5}
ggplot(model_dat, aes(x = mag, fill = epoch)) +
  geom_density(alpha = 0.45) +
  scale_x_reverse() +
  scale_fill_manual(values = c(pre_2022 = "steelblue",
                               dimming_2023on = "firebrick")) +
  labs(title = "Magnitude distribution: pre-2022 vs. 2023-onward",
       x = "Magnitude (fainter →)", y = "Density", fill = "Epoch") +
  theme_minimal(base_size = 12)
```

\newpage

# Model 1 — Mean brightness (LMM)

We first model the **mean** magnitude. Magnitude is a continuous measurement, so
this is a linear mixed model (Gaussian response), not a count GLM.

- **Fixed effect:** `epoch` — the pre-vs-post contrast we care about.
- **Random effect:** `(1 | band)` — each passband gets its own intercept,
  absorbing the systematic zero-point differences between filters.

$$\text{mag}_{ij} = \beta_0 + \beta_1\,\text{epoch}_i + b_j^{\text{band}} + \varepsilon_{ij}$$

```{r lmm-fit}
m_mean <- lmer(mag ~ epoch + (1 | band), data = model_dat, REML = TRUE)
summary(m_mean)
```

```{r lmm-effect}
# The epoch coefficient = change in mean magnitude, holding band constant.
tidy(m_mean, effects = "fixed", conf.int = TRUE) |>
  kable(digits = 4,
        caption = "Fixed effects. 'epochdimming_2023on' > 0 means fainter since 2023.")
```

# Model 2 — Variability (location–scale GLMM)

The central question is about **variability**, not just the average. A standard
mixed model assumes one common residual variance. Here we let the residual
**dispersion depend on epoch** via `glmmTMB`'s `dispformula`, so the model
estimates a separate spread for each era and tests whether they differ.

```{r ls-fit}
m_scale <- glmmTMB(
  mag ~ epoch + (1 | band),
  dispformula = ~ epoch,          # <- variance is allowed to change by epoch
  data   = model_dat,
  family = gaussian()
)
summary(m_scale)
```

```{r ls-interpret}
# Back-transform the dispersion model to a residual SD per epoch.
disp <- fixef(m_scale)$disp
sd_pre     <- sqrt(exp(disp[["(Intercept)"]]))
sd_dimming <- sqrt(exp(disp[["(Intercept)"]] + disp[["epochdimming_2023on"]]))

tibble(
  epoch          = c("pre_2022", "dimming_2023on"),
  residual_sd    = c(sd_pre, sd_dimming)
) |>
  kable(digits = 3,
        caption = "Model-estimated residual SD by epoch (the variability answer)")

cat(sprintf(
  "Variability change: residual SD goes from %.3f (pre-2022) to %.3f (2023 on) — a %.0f%% %s.\n",
  sd_pre, sd_dimming,
  100 * abs(sd_dimming - sd_pre) / sd_pre,
  ifelse(sd_dimming > sd_pre, "increase", "decrease")
))
```

## Cross-check: simple variance test

An independent, model-free check on the same claim.

```{r var-test}
var.test(mag ~ epoch, data = model_dat) |>
  broom::tidy() |>
  kable(digits = 4, caption = "F test for equal variances between epochs")
```

\newpage

# Diagnostics

```{r diagnostics, fig.width=10, fig.height=4}
diag_df <- tibble(
  fitted   = fitted(m_mean),
  residual = resid(m_mean)
)

p1 <- ggplot(diag_df, aes(fitted, residual)) +
  geom_point(alpha = 0.3, size = 1, colour = "steelblue") +
  geom_hline(yintercept = 0, colour = "firebrick") +
  geom_smooth(method = "loess", se = FALSE,
              colour = "grey30", linetype = "dashed") +
  labs(title = "Residuals vs Fitted", x = "Fitted", y = "Residual") +
  theme_minimal(base_size = 11)

p2 <- ggplot(diag_df, aes(sample = residual)) +
  stat_qq(alpha = 0.3, size = 1, colour = "steelblue") +
  stat_qq_line(colour = "firebrick") +
  labs(title = "Normal Q-Q", x = "Theoretical", y = "Sample") +
  theme_minimal(base_size = 11)

p1 + p2
```

# +/-3 SD summary by epoch

```{r sd-summary}
model_dat |>
  group_by(epoch) |>
  group_modify(~ sd3(.x$mag)) |>
  pivot_wider(names_from = epoch, values_from = value) |>
  kable(digits = 3, caption = "Magnitude +/-3 SD summary by epoch")
```

\newpage

# Sensitivity check — Johnson V band only {#sec-vonly}

The pooled models above adjust for band with a random intercept, but they still
pool the **residual scatter** across bands whose composition shifts over time
(recent years are `CR`-heavy). To check whether the variability conclusion is an
artifact of that shift, we repeat the analysis on the single most homogeneous,
well-populated passband — **Johnson `V`** — where a like-for-like comparison
needs no band effect at all.

```{r vonly-data}
v_dat <- model_dat |> filter(band == "V") |> droplevels()

v_dat |>
  group_by(epoch) |>
  summarise(n = n(), mean_mag = mean(mag), sd_mag = sd(mag), .groups = "drop") |>
  kable(digits = 3, caption = "V band only: brightness and variability by epoch")
```

With a single band there is no random effect to include, so we fit a
location–scale model with `epoch` in **both** the mean and the dispersion:

```{r vonly-fit}
m_v <- glmmTMB(
  mag ~ epoch,
  dispformula = ~ epoch,
  data   = v_dat,
  family = gaussian()
)
summary(m_v)

# Mean shift and per-epoch residual SD, back-transformed.
dv <- fixef(m_v)$disp
tibble(
  quantity = c("mean shift (mag, + = fainter)",
               "residual SD pre_2022",
               "residual SD dimming_2023on"),
  value = c(
    fixef(m_v)$cond[["epochdimming_2023on"]],
    sqrt(exp(dv[["(Intercept)"]])),
    sqrt(exp(dv[["(Intercept)"]] + dv[["epochdimming_2023on"]]))
  )
) |>
  kable(digits = 3, caption = "V-only: mean and variability change")
```

```{r vonly-vartest}
var.test(mag ~ epoch, data = v_dat) |>
  broom::tidy() |>
  kable(digits = 4,
        caption = "V-only F test for equal variances between epochs")
```

**Interpretation.** Restricted to a single band, the dimming (~1.7 mag fainter)
persists **and** the variability drop that the pooled model washed out
reappears clearly (residual SD roughly halves). This tells us the flat residual
SD in the pooled model was driven by changing band composition, not by the
source: measured like-for-like, 3C 279 has become both **fainter and quieter**
since 2023.

\newpage

# Summary

- **Data:** AAVSO magnitudes for 3C 279, cleaned to trusted photometric bands
  (`r paste(KEEP_BANDS, collapse = ", ")`), spanning ~1999 to present.
- **Mean brightness:** Model 1 estimates the average magnitude shift from the
  pre-2022 era to the 2023-onward era (see the `epoch` coefficient — positive =
  dimmer). The object is clearly fainter post-2023.
- **Variability:** Model 2 (location–scale GLMM) estimates a **separate spread
  for each epoch**. On the *pooled* data the residual SD looks nearly unchanged,
  but that is a band-composition artifact — see the caveat below.
- **Sensitivity check (@sec-vonly):** restricting to Johnson `V` alone removes
  the band-mix confound and shows the variability genuinely **dropped** after
  2023 (residual SD roughly halves), alongside the dimming. The like-for-like
  conclusion is: 3C 279 became both fainter and quieter.
- **Caveats:** epochs are unbalanced across bands (e.g. `CR` dominates recent
  years); the `(1 | band)` term adjusts for filter zero-points but not for
  changing band *composition* over time — which is exactly why the single-band
  check matters. A useful next step is adding an observer effect if the raw
  AAVSO download includes an observer code.
```

