---
title: "GLMM: Fixed and Random Effects on Wolf Numbers"
author: "AAVSO Solar Section"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
# Packages -------------------------------------------------------------------
library(tidyverse)
library(hglm)
library(MASS)
library(scales)    # for plot formatting
library(patchwork) # for combining plots

# Project root ---------------------------------------------------------------
WD <- "C:/Users/Owner/Documents/SSN3"

# Reporting period: previous calendar month ----------------------------------
today  <- Sys.Date()
yr_num <- as.integer(format(today, "%Y"))
mo_num <- as.integer(format(today, "%m")) - 1L

if (mo_num == 0L) {          # roll back across December
  yr_num <- yr_num - 1L
  mo_num <- 12L
}

yr  <- as.character(yr_num)
mo  <- sprintf("%02d", mo_num)   # zero-padded month
Ex  <- paste0(yr, mo)            # e.g. "202602"
ver <- "00"

# Solar Cycle 25 start date (December 2019, SIDC/NASA/NOAA declaration) ------
# Edit CYCLE_START_YEAR and CYCLE_START_MONTH to change the filter window.
CYCLE_START_YEAR  <- 2019L
CYCLE_START_MONTH <- 12L

cat("Reporting period:", Ex, "| version:", ver, "\n")
cat("Data filtered from SC25 start: ",
    CYCLE_START_YEAR, "/", sprintf("%02d", CYCLE_START_MONTH), "\n", sep = "")
```

```{r utilities}
# Utility functions ----------------------------------------------------------

#' Standard error
se <- function(x) sqrt(var(x, na.rm = TRUE) / sum(!is.na(x)))

#' Confidence interval around the mean
ci_mean <- function(x, conf = 0.95) {
  n      <- sum(!is.na(x))
  margin <- qt(conf + (1 - conf) / 2, df = n - 1) * se(x)
  tibble(
    mean  = mean(x, na.rm = TRUE),
    lower = mean(x, na.rm = TRUE) - margin,
    upper = mean(x, na.rm = TRUE) + margin,
    conf  = conf
  )
}

#' +/-3 SD summary
sd3 <- function(x) {
  m <- round(mean(x, na.rm = TRUE), 3)
  s <- round(sd(x,   na.rm = TRUE), 3)
  tibble(
    label = c("-3sd", "-2sd", "-1sd", "Mean", "+1sd", "+2sd", "+3sd"),
    value = c(m - 3*s, m - 2*s, m - s, m, m + s, m + 2*s, m + 3*s)
  )
}

#' Read a CSV from the project Data/ folder
fetch <- function(filename) {
  path <- file.path(WD, "Data", paste0(filename, ".csv"))
  read_csv(path, show_col_types = FALSE)
}

#' Write a CSV to the project Data/ folder
write_csv_out <- function(df, filename) {
  path <- file.path(WD, "Data", paste0(filename, ".csv"))
  write_csv(df, path)
  invisible(path)
}
```

```{r load-data}
# Load data ------------------------------------------------------------------
load(file.path(WD, "ymtd.RData"))   # loads as X

X_full <- as_tibble(X)   # retain full history if needed later

# Filter to Solar Cycle 25 start ---------------------------------------------
# Requires columns named 'year' and 'mon' (integer); adjust if named differently.
X <- X_full |>
  filter(
    year > CYCLE_START_YEAR |
    (year == CYCLE_START_YEAR & mon >= CYCLE_START_MONTH)
  )

n_dropped <- nrow(X_full) - nrow(X)
cat(sprintf(
  "Rows loaded: %d total | %d pre-SC25 dropped | %d retained\n",
  nrow(X_full), n_dropped, nrow(X)
))

glimpse(X)
summary(X)
```

\newpage

# Wolf Number Distribution

Folks even re-define the solar dynamo as the "solar alternator":
<https://journal.geophysicsjournal.com/JofG/article/view/320/260#nav-menu>

## Histogram with Theoretical Overlays

```{r wolf-histogram, fig.width=7, fig.height=5}
# Wolf number histogram ------------------------------------------------------

w_mean <- mean(X$w, na.rm = TRUE)
w_sd   <- sd(X$w,   na.rm = TRUE)
w_var  <- var(X$w,  na.rm = TRUE)

cat(sprintf("Wolf number  mean = %.3f   variance = %.3f\n", w_mean, w_var))

# Freedman-Diaconis bin width
bwidth <- diff(range(X$w, na.rm = TRUE)) /
          nclass.FD(na.omit(X$w))

plot_title <- paste0(Ex, " v", ver, " Wolf Number Distribution")

p_hist <- ggplot(X, aes(x = w)) +
  geom_histogram(
    aes(y = after_stat(density)),
    binwidth  = bwidth,
    fill      = "grey80",
    colour    = "grey45",
    linewidth = 0.3
  ) +
  # Normal overlay
  stat_function(
    aes(colour = "Normal"),
    fun       = dnorm,
    args      = list(mean = w_mean, sd = w_sd),
    linewidth = 0.9
  ) +
  # Poisson overlay (integer support)
  geom_point(
    data = tibble(
      w    = 0:max(X$w, na.rm = TRUE),
      dens = dpois(0:max(X$w, na.rm = TRUE), lambda = w_mean)
    ),
    aes(x = w, y = dens, colour = "Poisson"),
    size = 1.2
  ) +
  scale_colour_manual(
    name   = "Theoretical density",
    values = c(Normal = "black", Poisson = "firebrick")
  ) +
  labs(
    title = plot_title,
    x     = "Wolf Number",
    y     = "Density"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(hjust = 0.5, face = "bold"),
    legend.position = "bottom"
  )

p_hist

# Save plot
ggsave(
  filename = file.path(WD, "Plots", paste0(Ex, ver, "HistW.png")),
  plot     = p_hist,
  width    = 7,
  height   = 5,
  dpi      = 150
)
```

\newpage

# HGLM: Hierarchical Generalised Linear Model

## Model Fitting

Wolf numbers are overdispersed counts. A **quasipoisson** fixed-effects family with a **Gamma**
random-effects family accounts for both the count nature of the data and between-observer
heterogeneity.

- **Fixed effects:** `see`, `sidc`, `year`, `mon` -- known reference counts and temporal structure
- **Random effect:** `(1 | obs)` -- observer-level intercept capturing systematic reporting differences

```{r hglm-fit, results='asis'}
part <- "hglm"

m1 <- hglm(
  fixed       = w ~ see + sidc + year + mon,
  random      = ~ 1 | obs,
  family      = quasipoisson(link = "log"),
  rand.family = Gamma(link = "log"),
  data        = X
)

print(summary(m1), print.ranef = TRUE)

# Save model -----------------------------------------------------------------
model_path <- file.path(WD, "models", paste0(Ex, ver, part, ".RData"))
m <- m1
save(m, file = model_path, ascii = FALSE)
cat("\nModel saved to:", model_path, "\n")
```

\newpage

# Exploratory Analysis of HGLM

```{r hglm-explore}
# Reload saved model ---------------------------------------------------------
load(model_path)

ms <- summary(m)
print(ms, print.ranef = TRUE)

# Flagged observations (hat-value = 1 candidates) ----------------------------
if (length(m$bad) > 0) {
  cat("Flagged observations (m$bad):", m$bad, "\n")
} else {
  cat("No flagged observations.\n")
}

# Fitted vs. observed --------------------------------------------------------
# hglm has no $frame; recover observed as fitted + working residual
diag_df <- tibble(
  fitted   = m$fv,
  residual = m$resid,
  observed = m$fv + m$resid
)

cat(sprintf(
  "\nFitted values -- mean: %.3f   variance: %.3f   ratio: %.3f\n",
  mean(diag_df$fitted),
  var(diag_df$fitted),
  var(diag_df$fitted) / mean(diag_df$fitted)
))
```

## Diagnostic Plots

```{r hglm-diagnostics, fig.width=10, fig.height=4}
p1 <- ggplot(diag_df, aes(x = fitted, y = residual)) +
  geom_point(alpha = 0.3, size = 1, colour = "steelblue") +
  geom_hline(yintercept = 0, colour = "firebrick", linewidth = 0.7) +
  geom_smooth(method = "loess", se = FALSE,
              colour = "grey30", linewidth = 0.7, linetype = "dashed") +
  labs(title = "Residuals vs Fitted",
       x = "Fitted values", y = "Residual") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(hjust = 0.5))

p2 <- ggplot(diag_df, aes(sample = residual)) +
  stat_qq(alpha = 0.3, size = 1, colour = "steelblue") +
  stat_qq_line(colour = "firebrick", linewidth = 0.7) +
  labs(title = "Normal Q-Q of Residuals",
       x = "Theoretical quantiles", y = "Sample quantiles") +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(hjust = 0.5))

p1 + p2

# Save diagnostics
ggsave(
  filename = file.path(WD, "Plots", paste0(Ex, ver, part, "Diag.png")),
  plot     = p1 + p2,
  width    = 10,
  height   = 4,
  dpi      = 150
)
```

## SD Summary of Wolf Numbers

```{r sd-summary}
sd3(X$w) |>
  knitr::kable(
    digits    = 3,
    caption   = "Wolf Number +/-3 SD Summary",
    col.names = c("Label", "Value")
  )
```
